Improving the Accuracy of a Sequential Mining Model for Hurricane Trajectory Prediction

### 2018-01-29 paper/manuscript_revised_jan25.tex
Keep paper to 14 pages

In HTPDM and WARD-HTP, the algorithm that generates trajectory prediction possibilities is based on AprioriAll. AprioriAll was first proposed at IBM research as a method for discovering frequent sequential patterns in transactional databases \cite{AprioriAll-original}. In its original implementation, AprioriAll is able to generate association rules describing sentences of the form ``Customers who buy P may later buy Q''. In the hurricane trajectory prediction problem, AprioriAll is used to generate association rules describing sentences of the form ``Hurricanes located at P may later be located at Q''. In the context of AprioriAll, hurricanes are considered $customers$ and their sequences of recorded points are considered $transactions$. Consider the high-level implementation steps of AprioriAll as follows:

@@

Section (2) of AprioriAll is responsible for the generation of frequent sequences. In the case of hurricane trajectory prediction, these sequences correspond to frequently recorded hurricane coordinate sequences. Frequent hurricane coordinate sequences are those which occur at least a minimum number of times, as defined by the \textit{minimum support}, or minsup. In general terms, the sequence generation step of AprioriAll develops frequent ordered patterns of increasing length, beginning with frequent single-item sequences (corresponding to frequent single-point sequences). This step begins by scanning the $\langle HurricaneID, CoordinateSequence\rangle$ database for frequent single-point sequences. The frequent single-point trajectories are then converted to candidate two-point trajectories as the result of an inner join. The candidate two-point trajectories are scanned against the database to determine which candidates satisfy the minimum support. Another inner join generates the candidate 3-length trajectories. Once the sequence generation level reaches three, each candidate sequence must be pruned if any of their candidate subsequences are infrequent. Continuous execution of AprioriAll until termination results in the collection of all frequent hurricane trajectory patterns defined by the minimum support.

@@

Figure (7) shows the correctness ratios generated by all possible training data and testing data pairings. Note that ratios in brackets denote the worse-case ratio while ratios outside of brackets refer to the best-case. Between all possible options, we find that the training data interval of 1950-2000 results in the best correctness ratios for testing data from 2001-2015 when applying all other parameters we deemed optimal throughout the previous results of this work. Regardless of correctness ratio, 2001-2015 should always be selected as the test data interval to encourage analysis against more recent hurricane patterns. The results show that no training-data/testing-data pairing was able to generate correctness ratios higher than the ones generated by training the model with data from 1950 to 2000, and testing it on data as recently recorded as 2015.

@@

This work proposed a new model named the Weighted-Asset, Realistic-Distance Hurricane Trajectory Predictor (WARD-HTP) for improving the accuracy of existing data mining approaches to hurricane trajectory prediction. An AprioriAll-based hurricane trajectory prediction model similar to \cite{major-hurricane-model} was developed to demonstrate the unique contributions of WARD-HTP. There are three such contributions brought forth in the development and implementation of the model proposed. These contributions are the use of training data from 1950 to 2000, the use of weighted training data, or weighted assets (WA), and the use of the Haversine Distance Formula for realistic distance (RD). The three contributions combined form WARD-HTP and produce a hurricane trajectory prediction correctness ratio of exactly 75.0\%, which is 10\% higher than the current state-of-the-art. When accounting for trajectories which were not able to match with any association rules, the correctness ratio of WARD-HTP is 72.58\%, which is still 5.08\% higher than the worst case state-of-the-art ratio of 57.5\% as seen in \cite{major-hurricane-model}.

There are multiple areas to explore in the hurricane trajectory prediction problem space with respect to future work. Experiments in this work show that all datasets contain a subset of trajectories for which no association rule could be matched. While \cite{major-hurricane-model} proposed an extrapolation method to recover from rule-miss incidents, more work must be completed to reduce the number of non-matchable trajectories. Many of the mismatched trajectories are outliers with very little conformance to existing hurricane trajectories as seen in figure (3). Additionally, more experiments will be run on different inputs including training data from other regions such as the Pacific region. Further approaches to increasing the correctness ratio of WARD-HTP will also be investigated. This includes the investigation of alternative methods for defining training data weight.

>>>>

In HTPDM and WARD-HTP, the algorithm that generates trajectory prediction possibilities is based on AprioriAll. In its original implementation, AprioriAll is able to generate association rules describing sentences of the form ``Customers who buy P may later buy Q'' \cite{AprioriAll-original}. In the hurricane trajectory prediction problem, AprioriAll is used to generate association rules describing sentences of the form ``Hurricanes located at P may later be located at Q''. In the context of AprioriAll, hurricanes are considered $customers$ and their sequences of recorded points are considered $transactions$. Consider the high-level implementation steps of AprioriAll as follows:

@@

Section (2) of AprioriAll is responsible for the generation of frequent sequences. In the case of hurricane trajectory prediction, these sequences correspond to frequently recorded hurricane coordinate sequences. Frequent hurricane coordinate sequences are those which have a support (occurance count) greater than or equal to the \textit{minimum support}. In general terms, the sequence generation step of AprioriAll develops frequent ordered patterns of increasing length, beginning with frequent single-item sequences (corresponding to frequent single-point sequences). This step begins by scanning the $\langle HurricaneID, CoordinateSequence\rangle$ database for frequent single-point trajectories. The frequent single-point trajectories are then converted to candidate two-point trajectories by an inner join. The candidate two-point trajectories are scanned against the database to determine which candidates satisfy the minimum support. Running an inner join against the frequent k-point trajectories generates the candidate (k+1)-point trajectories. Once the sequence generation level reaches 3, each candidate sequence must be pruned if any of their subsequences are infrequent. The complete exectuion of AprioriAll results in a collection of all frequent hurricane trajectory patterns defined by the minimum support.

@@

Figure (7) shows the correctness ratios generated by all pairings of training and testing data. Note that ratios in brackets denote the worse-case ratio while unbracketed ratios refer to the best-case. Between all possible options, we find that the training data interval of 1950-2000 results in the best correctness ratios for testing data from 2001-2015 when applying all other parameters we deemed optimal throughout the previous results of this work. Regardless of correctness ratio, 2001-2015 should always be selected as the test data interval to encourage analysis against more recent hurricane patterns. The results show that no training-data/testing-data pairing was able to generate correctness ratios higher than those generated by training the model with data from 1950 to 2000, and testing it on data from 2001 to 2015.

@@

This work proposed a new model named the Weighted-Asset, Realistic-Distance Hurricane Trajectory Predictor (WARD-HTP) for improving the accuracy of existing data mining approaches to hurricane trajectory prediction. An AprioriAll-based hurricane trajectory prediction model similar to \cite{major-hurricane-model} was developed to demonstrate the unique contributions of WARD-HTP. There are three such contributions brought forth in the development and implementation of the model proposed. These contributions are the use of training data from 1950 to 2000, the use of weighted training data, and the use of the Haversine Distance Formula.. The three contributions combined form WARD-HTP and produce a hurricane trajectory prediction correctness ratio of exactly 75.0\%, which is 10\% higher than the current state-of-the-art. When accounting for trajectories which were not able to match with any association rules, the correctness ratio of WARD-HTP is 72.58\%, which is still 5.08\% higher than the worst case state-of-the-art ratio of 57.5\% as seen in \cite{major-hurricane-model}.

There are multiple aspects of future work to explore in the hurricane trajectory prediction problem space. Experiments in this work show that all datasets contain a subset of trajectories for which no association rule could be matched. While \cite{major-hurricane-model} proposed an extrapolation method to recover from rule-miss incidents, more work must be completed to reduce the number of non-matchable trajectories. Many of the mismatched trajectories are outliers with very little conformance to existing hurricane trajectories as seen in figure (3). Additionally, more experiments will be run on different inputs including training data from other regions such as the Pacific region. Further approaches to increasing the correctness ratio of WARD-HTP will also be investigated, including the pursuit of alternative methods for implementing AprioriAll with weighted transactions.
